---
title: "Homework 2"
author: "Anna Ziff"
date: "R Workflow for Economists"
output: pdf_document
header-includes:
    - \usepackage{fancyhdr}
    - \usepackage{hyperref}
    - \pagestyle{fancy}
    - \fancyfoot[R]{Anna Ziff}
    - \fancyfoot[L]{}
    - \fancyhead[R, C, L]{}
    - \renewcommand{\headrulewidth}{0pt}
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. What is the difference between \texttt{geom\_line()} and \texttt{geom\_path()}? Demonstrate the difference with this example tibble.
```{r, eval = FALSE}
toy <- tibble(
    x = c(3, 1, 5),
    y = c(2, 4, 6),
    label = c("a", "b", "c")
)
```

2. Load the dataset \texttt{gapminder\_large.csv}. Reshape the dataset so that it is in long rather than wide format.

3. Create a plot that graphs the average CO2 emissions by region across time. Feel free to add any elements and formatting you like. Create a table with summary statistics describing per capita CO2 emissions by region and year. Format both the graph and the table and describe them as you would in an academic paper.

4. Refer to [3.8 Exercises](https://link.springer.com/content/pdf/10.1007/978-0-387-77318-6_3.pdf), page 91. Complete exercise 1 and one other exercise of your choice from that list that aligns with your interests. You may need to use the \texttt{data()} function to download the data. For example, the data for exercise 2 can be obtained as follows.
```{r, eval = FALSE}
library(AER)

# Load data
data("HousePrices", package = "AER")

# Description of data
?HousePrices
```



